---
title: "Reproducible Data: Peer Review Assignment 1"
author: "Clyde Tressler"
date: "November 5, 2015"
output: html_document
---
```{r global_options, echo=TRUE}
```
<div style="text-align: center"><h1 >Reproducible Data: Peer Review Assignment 1</h1>
<h4>Clyde Tressler </br>November 5, 2015</h4>
</div>
<h3> Introduction</h3>
In this report we will demonstrate the principles of reproducible analyses by providing an integrated document that combines source code and code output. We use 2 months of observations from a personal activity monitoring device.

<h3>Reading the Data</h3>
First we load the activity monitoring data from the csv file. We use strings as character vectors to facilitate conversion of the date column to POSIX dates. 

```{r load_data}
setwd('~/RepData_PeerAssessment1')
dat <- read.csv('activity.csv', stringsAsFactors = FALSE)
dat$date <- as.Date(dat$date)
str(dat)
```
<h3>Analysis of Raw Data</h3>
Below we group interval observations by day, then plot a histogram of steps by day and report the mean and median number of steps per day. NA observations are removed when summing the interval data. The dplyr library is used to group and summarize the intervals by date.
```{r, message=FALSE}
library(dplyr)
```

```{r plot_histogram_1}
date_grouped <- group_by(dat, date)
steps <- summarise_each(date_grouped, funs(sum(., na.rm = TRUE)), steps)
head(steps)
hist(steps$steps, main = 'Histogram of Steps per Day', xlab = 'Number of Steps', ylim=c(0,20),
     col='steelblue', breaks=10)
est_mean <- mean(steps$steps)
est_median <- median(steps$steps)
```

```{r report_mean_median}
est_mean
est_median
```

Next we group the data by 5-minute interval values in order to visualize the average steps per interval over all days in the observations. 

```{r plot_time_series}
interval_grouped <- group_by(dat, interval)
interval_steps <- summarise_each(interval_grouped, funs(mean(., na.rm = TRUE)), steps)
head(interval_steps)
tail(interval_steps)
plot(interval_steps$interval, interval_steps$steps, main = 'Number of Steps per 5-minute Interval',
     xlab = 'Interval', col='steelblue', type = 'l', ylab = 'Number of Steps', xlim=c(0,2500))
```

We verify that the number of intervals observed corresponds to 24 hours and then identify the interval with the maximum number of steps:
```{r calc_max}
length(unique((dat$interval)))/12
max_steps <- subset(interval_steps, steps==max(steps))
```
Maximum number of steps is `r round(max_steps[1,2], 2)` during interval `r max_steps[1,1]`.  

Next we calculate the number of NA observations in the dataset:
```{r}
sum(is.na(dat$steps))
```
<h3>Imputing Values for Missing Observations</h3>
Now we use vector recycling in R to quickly assign the step count means for each interval as a column in the original data frame, and then add another column to impute the missing values of the step observations by substituting the mean value when an NA is present in the original data.

```{r imputing_nas}
unique(dat$interval == interval_steps$interval) #check to see the intervals all match
dat$int_means <- interval_steps$steps
dat$imputed_steps <- ifelse(is.na(dat$steps), dat$int_means, dat$steps)
```
Let's make sure it worked:
```{r check_chunk}
head(dat)
tail(dat)
dat[14000:14006,] #looks right!
```

Now we repeat the analyses using the imputed data:
```{r plot_histogram_2}
date_grouped <- group_by(dat, date)
imputed_steps <- summarise_each(date_grouped, funs(sum(.)), imputed_steps)
head(imputed_steps)
hist(imputed_steps$imputed_steps, main = 'Histogram of Steps per Day',
     xlab = 'Number of Steps', ylim=c(0,25), col='steelblue', breaks=10)
mean(imputed_steps$imputed_steps)
median(imputed_steps$imputed_steps)
```

Are the values for the mean and median with the missing values filled-in the same as the previous calculations?
```{r examine_changes}
est_mean == mean(imputed_steps$imputed_steps)
est_median == median(imputed_steps$imputed_steps)
```
No, the values are different. In this case, replacing the missing data has resulted in step counts that seem to be more normally distributed around the mean.

<h3>Weekday Activity vs Weekend Activity</h3>
Now we examine whether activity patterns are different on weekends versus weekdays. We add a column using a factor variable to distinguish weekdays from weekends and a 'day of week' column to visually verify the factor values. We provide panel plots of the interval data for weekends and weekdays:
```{r plot_day_of_week}
weekends <- c('Saturday', 'Sunday')
dat$dow <- weekdays(dat$date)
dat$day <- factor((weekdays(dat$date) %in% weekends), levels=c(TRUE, FALSE), labels=c('weekend', 'weekday'))
interval_grouped <- group_by(dat, interval, day)
interval_steps <- summarise_each(interval_grouped, funs(mean(.)), imputed_steps)
library(lattice)
xyplot(interval_steps$imputed_steps ~ interval_steps$interval | day, data=interval_steps,
       layout=c(1,2), main="Mean Steps for 5-Minute Intervals for Weekdays vs Weekends",
       xlab="Interval", ylab="Number of Steps", type=c("l","l"))
```

Now we use a smoothing filter to better understand the important features of the plots:
```{r smoothed_plot, fig.width=4}
xyplot(interval_steps$imputed_steps ~ interval_steps$interval | day, data=interval_steps,
       layout=c(1,2), main="Mean Steps with Smoothed Intervals", xlab="Interval",
       ylab="Number of Steps", type=c("smooth","smooth"), span=.25, ylim = c(0,100))
```

<h3>Interpretation</h3>
The plots show that the weekday step counts peak early in the day, plateau for some hours, and then increase later in the day. The weekend step counts build quickly and do not plateau. This seems consistent with a traditional 40-hour work week schedule.

